Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-15 index.qmd
section:issues; process:write

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp. 254] (see @fig-thurstone_cases). This substitution has little effect on the model's estimation or interpretation, as the Normal and logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (refer to @fig-logistic_vs_normal).

@@

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement. As a result, its assumptions may not be suitable for applications beyond the psychophysical contexts, which typically involve simple, homogeneous stimuli. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], further acknowledging that some assumptions could become problematic with less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretical or in practice: the assumption of zero correlation and equal dispersion between stimuli.

>>>>

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp. 254] (see @fig-thurstone_cases). This substitution has little effect on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (refer to @fig-logistic_vs_normal).

@@

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement. As a result, its assumptions may not be suitable for applications beyond the psychophysical contexts, which typically involve simple, homogeneous stimuli. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], further acknowledging that some assumptions could become problematic with less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretically or in practice: the assumption of zero correlation and equal dispersion between stimuli.
